---
title: "Discovering disproportionate jail rate of black people in the United States"
author: "Victor Ahn"
date: '02.25.2022'
output: html_document
---
### Introduction  
  
  The United States is known for its _high incarceration rate_ within the country. There have been unsolved problems and questions over the _disproportionate incarceration rate of people of color_ compared to white people. This project will focus on _**how the jail rate (People in jail /Population) varies between white people and black people in different regions**_ in the United States (South, West, Midwest) since the start of the 21st century. 
```{r setup, include=FALSE}
## collection of R packages for D.S
library(tidyverse) 
## A new package which provides a set of tools for efficiently manipulating datasets in R
library(dplyr)
## A R package dedicated to Data Visualization
library(ggplot2)
## A package for creating interactive, publication-quality graphs
library(plotly)
## A package that contains a lot of outlines of continents, countries, states, and counties that have been with R
library(maps)
## A package that contains demographic data on the United States at the county and state levels
library(usdata)

## Data Loading
incarceration <- read.csv("https://raw.githubusercontent.com/vera-institute/incarceration-trends/master/incarceration_trends.csv")

## Filtered Data for jail rates data
entire_country_jail_rate <- incarceration %>% mutate(total_jail_rate = round(((total_jail_pop / total_pop)*100),2)) %>% 
  mutate(black_jail_rate = round(((black_jail_pop / black_pop_15to64)*100),2)) %>% mutate(white_jail_rate = round(((white_jail_pop / white_pop_15to64)*100),2)) %>% 
  select (year, region, total_jail_rate, white_jail_rate, black_jail_rate)

```
  
  
### Summary Information

  Values that I found relevant are _mean of total jail rates in three different regions_, mean of white jail rates in three different regions, mean of black jail rates in three different regions and mean of black jail rate over white jail rate in three different regions. First of all, it is **important** to use the jail rate instead of jail population because the total population varies among race groups. Absolute numbers cannot testify whether disproportionate incarceration is true. _Comparing relative percentages will be a right way to find out the existence of disproportion._ These values I chose are necessary because mean values are helpful to compare jail rates between white and black in different regions in the U.S during the 21st century. Mean value of black jail rate over white jail rate is also important because it demonstrates how many times more certain groups are incarcerated than the other.
  
```{r, include=FALSE}
## Data of jail rates in south region
south_states <- entire_country_jail_rate %>% filter(region == "South") %>% slice(31:49)
mean_south_total <- round(mean(south_states$total_jail_rate),2)
mean_south_white <- round(mean(south_states$white_jail_rate),2)
mean_south_black <- round(mean(south_states$black_jail_rate),2)

## Black jail rate over white jail rate data in south region
black_white_diff_south <- south_states %>% mutate(black_white_diff = round((black_jail_rate / white_jail_rate),2)) %>% 
                                                    select(year, region, black_white_diff)

## Data of jail rates in west region
west_states <- entire_country_jail_rate %>% filter(region == "West") %>% slice (31:49)
mean_west_total <- round(mean(west_states$total_jail_rate),2)
mean_west_white <- round(mean(west_states$white_jail_rate),2)
mean_west_black <- round(mean(west_states$black_jail_rate),2)

## Black Jail Rate / White Jail Rate (West)
black_white_diff_west <- west_states %>% mutate(black_white_diff = round((black_jail_rate / white_jail_rate),2)) %>% 
  select(year, region, black_white_diff)

## Data of jail rates in south region
midwest_states <- entire_country_jail_rate %>% filter(region == "Midwest") %>% slice(31:49)
mean_midwest_total <- round(mean(midwest_states$total_jail_rate),2)
mean_midwest_white <- round(mean(midwest_states$white_jail_rate),2)
mean_midwest_black <- round(mean(midwest_states$black_jail_rate),2)

## Black Jail Rate / White Jail Rate (Midwest)
black_white_diff_midwest <- midwest_states %>% mutate(black_white_diff = round((black_jail_rate / white_jail_rate),2)) %>% 
  select(year, region, black_white_diff)

## Mean value of black jail rate over white jail rate in different regions
mean_south <- round(mean(black_white_diff_south$black_white_diff),2)
mean_west <- round(mean(black_white_diff_west$black_white_diff),2)
mean_midwest <- round(mean(black_white_diff_midwest$black_white_diff),2)
```
  
  **Here are the results and interpretations of my values:**
  
Mean of total population jail rate (South/West/Midwest):
**0.29% / 0.1% / 0.13%**
_(Average total jail rate during 2000 to 2018 is the highest in the South, then Midwest, and lowest in the West.)_
 
Mean of black jail rate (South/West/Midwest):
**0.93% / 0.18% / 1.13 %**
_(Average jail rate of black people during 2000 to 2018 is the highest in the Midwest, then South, and lowest in the West.)_
 
Mean of white jail rate (South/West/Midwest):
**0.27% / 0.55% / 0.15%**
_(Average jail rate of white people during 2000 to 2018 is the highest in the West, then in the South and lowest in the Midwest.)_
 
Mean of black jail rate over white jail rate (South/West/Midwest):
**3.49 / 0.25 / 6.08**
_(Average jail rate of black people is 3.49 times higher than that of white people in the South, 6.08 times higher in the Midwest. It is 0.25 times higher in the West which means the jail rate of white people is 4 times higher in the west than that of black people.)_


### Black Over White Jail Rate during 2000-2018 by Regions 
```{r, include=FALSE}
## Combined data of black jail rate over white jail rate data in different regions
black_white_diff_combined <- rbind(black_white_diff_west, black_white_diff_south, black_white_diff_midwest)
ggp_total <- ggplot(black_white_diff_combined, aes(x = year, y = black_white_diff, color = region)) + geom_line()
ggp_total <- ggp_total + labs(title = "Black Over White Jail Rate By Regions (2000-2018)", y = "Black over White Jail Rate", x= "Year", color = "Region")
```

```{r}
ggp_total
```


**Black over white jail rat**es are values that signify _how much higher jail rates of black people are compared to jail rates of white people._ If the number is lower than 1, it means that the jail rates of white people were higher in that region. If the number is equal to 1, it means jail rates of black and white people are the same. If the number is higher than 1, the jail rates of black people surpass those of white people. According to the trend over time graph that I demonstrated, in Midwest and west regions, black over white jail rates, although the number is still higher than 1, are going down. This means that in those two regions, the disproportionate incarceration rates between black and white people are going down. On the other hand, in the west region, the black over white jail rates, even though they are the lowest among three regions, are going up. This means that the disproportionate incarceration rates between black and white people are going up. However, the absolute disproportionate incarceration rates between black and white people are still much higher in the Midwest and south region than the west region. 

### Change in Jail Rate of Black in Different Regions (2000-2018)
```{r, include = FALSE}
## Change in jail rate of black people in different regions during 2000-2018 (Variable Comparison Chart)
black_south <- south_states %>% select(year, region, black_jail_rate)
black_west <- west_states %>% select(year, region, black_jail_rate)
black_midwest <- midwest_states %>% select(year, region, black_jail_rate)
black <- rbind(black_south, black_west, black_midwest)

```

```{r, include = FALSE}
bar_black <- ggplot(black, aes(x = year, y = black_jail_rate, fill = region)) + geom_col(position = "dodge")
bar_black <- bar_black + labs(title = "Change in Jail Rate of Black in Different Regions (2000-2018)", y = "Black Jail Rate (%)", x = "Year", color = "Region")
```

```{r}
bar_black
```


This bar chart specifically looks into change in jail rates of black people in different regions (South, West and Midwest). The reason why I am presenting this bar chart is because the **bar chart demonstrates more clearly about how jail rates changed over time.** As the bar chart shows, the jail rates of black people in the Midwest have gone down dramatically. The jail rates in the South have also been going down slightly. On the other hand, the jail rates in the West have gone up. Bar chart also clarifies that, even though the rates have gone down in the South and gone up in the West, the absolute jail rates are still higher in the South than the West.

### U.S Black Jail Rate Over Total Jail Rate in 2018 in the United States (Map)
```{r, include = FALSE}
## Black jail rate over total jail rate in 2018 in the United States (MAP)
data_map <- incarceration %>% mutate(white_jail_rate = round(((white_jail_pop / white_pop_15to64)*100),2)) %>%  
  mutate(black_jail_rate = round((black_jail_pop / black_pop_15to64)*100),2) %>%
  select (year, state, county_name, white_jail_rate, black_jail_rate, fips) %>% filter(year == 2018)

data_map <- data_map %>% mutate(black_over_white = round(((black_jail_rate / white_jail_rate)*100),2)) %>% 
  select(state, county_name, black_over_white, fips)

map_shapes <- map_data("county") %>% 
  unite(polyname, region, subregion, sep = ",") %>% left_join(county.fips, by="polyname")

map_data <- map_shapes %>% left_join(data_map, by="fips")

blank_theme <- theme_bw() +
  theme(
    axis.line = element_blank(),
    axis.text = element_blank(),
    axis.ticks = element_blank(),
    axis.title = element_blank(),
    plot.background = element_blank(),
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank(),
    panel.border = element_blank()
  )

final_map <- ggplot(map_data) +
  geom_polygon(
    mapping = aes(x = long, y =lat, group = group, fill = black_over_white),
    color = "gray", size = 0.3
  ) + 
  coord_map() + 
  scale_fill_continuous(limits = c(0, max(map_data$black_over_white)), na.value = "white", low = "yellow", high = "red") + blank_theme +
  ggtitle("Black Jail Rate over White Jail Rate in the United States") +
  guides(fill=guide_legend(title="Black over white jail rate"))
```
```{r}
final_map
```


This map demonstrates how much higher jail rate of black people is compared to jail rate of white people in the entire country in 2018. Overall color of the map is yellow which means **the jail rates between the two races are not as significantly different in 2018.** There are some counties that stands out, but ,overall, the jail rates between the two are quite similar.This is a good sign because it means that _the U.S incarceration and jurisdiction system is not completely out of order._ However, it is important to not forget that there are still systemic racism going on in different areas in the United States and it is vital to find a way to fix the system. 



  
